---
title: "Practical 3: Cluster Detection"
subtitle: "EMBL Course: Data mining and integration with networks"
author: "Aaron Brooks"
date: "December 11, 2015"
output: 
  html_document:
    css: include/hint.css
    includes:
      in_header: include/hint.html
    toc: true
    toc_depth: 2
---

# Clustering Tutorial

## Load Data 

First we will read the kernel data from Jean-Karim. I have written a script to read and parse the results.

```{r echo = TRUE, message = FALSE, cache = FALSE}
devtools::source_url("https://raw.githubusercontent.com/scalefreegan/Teaching/master/DataIntegration/readData.R")
```

## Link-community detection

Link-community detection inverts the clustering problem. Rather than cluster similar nodes, it instead clusters similar edges. 

In link-community detection, nodes can now belong to several clusters. This is particularly useful in biological contexts, where one might like to express, for example, the different role a particular gene has in different contexts.

You can learn more about link-community detection in the [original paper](http://www.nature.com/nature/journal/v466/n7307/full/nature09182.html)

> This is a task.
>
> You should do 'X'.

<div id='hint'>
<h5 class="ui-closed" style="color:blue">Hint</h5>
```{r echo = TRUE, message = FALSE, cache = FALSE, id = "Mine"}
library(pheatmap)
library(plotly)
library(dplyr)
library(reshape2)
library(kernlab)
library(matrixcalc)
library(igraph)
library(linkcomm)

devtools::source_url("https://raw.githubusercontent.com/scalefreegan/Teaching/master/DataIntegration/readData.R")

data = data_reduced

head(data)
```
</div>

### Some in between stuff!

#### even H5!

<div id='hint'>
<h5 class="ui-closed" style="color:blue">Hint</h5>
```{r echo = TRUE, message = FALSE, cache = FALSE, id = "Mine"}
library(pheatmap)
library(plotly)
library(dplyr)
library(reshape2)
library(kernlab)
library(matrixcalc)
library(igraph)
library(linkcomm)

devtools::source_url("https://raw.githubusercontent.com/scalefreegan/Teaching/master/DataIntegration/readData.R")

data = data_reduced

head(data)
```
</div>
