---
title: "Data Integration Course: Data Exploration"
author: "Aaron Brooks"
date: "October 22, 2015"
output: html_document
runtime: shiny
---

```{r echo = FALSE, message = FALSE, cache = FALSE}
library(pheatmap)
library(plotly)
library(dplyr)
library(reshape2)
library(kernlab)
```

## Why?

Basic exploration of data sent by Jean-Karim for Data Integration course at EMBL Dec 2015

Data available at: https://oc.embl.de/index.php/s/qiOSCyvYRdxraRw

I wrote a script to process these data:

```{r echo = TRUE, message = FALSE, cache = FALSE}
devtools::source_url("https://raw.githubusercontent.com/scalefreegan/Teaching/master/DataIntegration/readData.R")
```

## What does the data look like?

For the examples, we will only use the reduced data set. It looks like this:

```{r}
data = data_reduced

head(data)
```

## Data set info

The data sets are: 

* __BP__: Pairwise gene semantic similarities (Resnik score) in the biological process domain
* __CODA80__: Predicted protein interactions from domain co-occurrence/fusion events
* __HIPPO__: Protein interactions transferred from model organisms to human by orthology
* __PI__: Protein interactions in human (from multiple databases)
* __TM__: Protein interactions derived from the [iHOP](http://www.ihop-net.org/UniPub/iHOP/help.html) text mining method

The data set is sparse. In fact, there are `r sum(complete.cases(data_reduced))` complete cases across the entire data set! 

### Below you can compute the number of complete cases for each data type.

```{r echo = FALSE}
checkboxGroupInput(inputId = "datatype1", label = "Select data type:", choices = colnames(data)[-c(1,2)], inline = T)
```

```{r echo = FALSE}
renderText({ paste("There are", sum(complete.cases(data[input$datatype1])), "complete cases across these data sets", sep= " " )})
```

## Visualize the data

```{r echo = FALSE}
selectInput(inputId = "datatype2", label = "Select data type:", choices = c(colnames(data_reduced)[-c(1,2)],"Combined"), selected = 1)
```

```{r echo = FALSE}
 renderPlotly({
   d = input$datatype2
   if (d == "Combined") {
     z = select(data, gene1, gene2)
     z$value = rowSums(data[,-c(1,2)],na.rm=T)
     g = sort(unique(c(z$gene1, z$gene2)))
     m = matrix(0, nrow = length(g),ncol = length(g), dimnames = list(g,g))
     m[cbind(z$gene1, z$gene2)] = z[,"value"]
     m[cbind(z$gene2, z$gene1)] = z[,"value"]
   } else {
     z = select(data, gene1, gene2, which(colnames(data)==d))
     z[is.na(z[,d]),d] = 0
     g = sort(unique(c(z$gene1, z$gene2)))
     m = matrix(0, nrow = length(g),ncol = length(g), dimnames = list(g,g))
     m[cbind(z$gene1, z$gene2)] = z[,d]
     m[cbind(z$gene2, z$gene1)] = z[,d]
   }
   rk = which(apply(m,1,function(i)sum(i)>0))
   ck = which(apply(m,2,function(i)sum(i)>0))
   k = intersect(rk,ck)
   m = m[k,k]
   m_clustx = hclust( dist( m ) )
   m_clusty = hclust( dist( t( m ) ) )
   x = m_clustx$labels[m_clustx$order]
   y = m_clusty$labels[m_clusty$order]
   p <- plot_ly(z = m[x,y], x = x, y = y, colorscale = "Greys", reversescale =T, zauto = F, zmin = 0, zmax = quantile(m[m>0],.9), type = "heatmap", filename="data")
   p
 })
```

